---
title: "Comparative assessment of filtration- and precipitation-based methods for the concentration of SARS-CoV-2 and other viruses from wastewater"
author: "Cameron Pellett"
date: "17/03/2022"
output: github_document
---

This document and github repo provides the r code for results (3.2) produced in 
*Comparative assessment of filtration- and precipitation-based methods for the concentration of SARS-CoV-2 and other viruses from wastewater*. The full report with code is included below. The source code for this document can be found in README.rmd. For a cleaner report with a html output, Chem-Con-Bangor-Bath.rmd can be used. qPCR recovery data can be found in Barbara_full_dataset.csv and some additional untidy notes and exploration can be found in notes.r. Results in section 3.1. can be found in another repo: https://github.com/CameronPellett/Spiked-virus-concentration-Bangor. Text may have been changed for final publication.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 9)
```

```{r packages}
library(tidyverse)
library(lme4)
library(ggrepel)
library(ggbeeswarm)
library(patchwork)
library(broom)

theme_set(theme_grey()+theme(panel.background = element_rect(fill = "grey96")))
```

```{r functions}
pvalue_fun <- function(p.value){
  if(is.na(p.value)){
    NA_character_
  }else if(p.value > 0.05){
    "> 0.05"
  }else if(p.value < 0.05 & p.value > 0.01){
    "< 0.05"
  }else if(p.value < 0.01 & p.value > 0.001){
    "< 0.01"
  }else if(p.value < 0.001){
    "< 0.001"
  }else {
    "something went wrong"
  }
}

pvalue_star <- function(p.value){
  if(is.na(p.value)){
    NA_character_
  }else if(p.value > 0.05){
    "."
  }else if(p.value < 0.05 & p.value > 0.01){
    "*"
  }else if(p.value < 0.01 & p.value > 0.001){
    "**"
  }else if(p.value < 0.001){
    "***"
  }else {
    "something went wrong"
  }
}

stat_paste_fun <- function(stat, log = FALSE){
  if(log != TRUE){
    paste0(str_remove_all(stat[["method"]][[1]], "\n\t"), ": ", 
           str_remove_all(names(stat[["statistic"]]), "Kruskal-Wallis "), " = ",
           round(stat[["statistic"]][[1]],1), ", ",
           if(length(stat[["parameter"]]) == 0)
           {""
           }else if(is.na(stat[["parameter"]])){
             ""
           }else if(length(stat[["parameter"]]) > 0){
             paste0("df = ", round(stat[["parameter"]], 0), ", ")
           } ,
           "p-value ",  pvalue_fun(stat[["p.value"]]))
  }else{
    paste0(str_remove_all(stat[["method"]][[1]], "\n\t"), " (log y): ", 
           str_remove_all(names(stat[["statistic"]]), "Kruskal-Wallis "), " = ",
           round(stat[["statistic"]][[1]],1), ", ",
           if(length(stat[["parameter"]]) == 0)
           {""
           }else if(is.na(stat[["parameter"]])){
             ""
           }else if(length(stat[["parameter"]]) > 0){
             paste0("df = ", round(stat[["parameter"]], 0), ", ")
           } ,
           "p-value ",  pvalue_fun(stat[["p.value"]]))
  }
}

plot_assumptions <- function(model, obs, bins = 30){
        #create tibble for use in ggplot of model assumptions
        df <- tibble(
                observations = obs,
                fitted = fitted(model),
                residuals = resid(model),
                standardised_residuals = sqrt(resid(model)^2),
                RMSresiduals = sqrt(mean(resid(model)^2)))
    
        
        a <- ggplot(df, aes(fitted, residuals))+
                geom_point()+
                geom_abline(aes(slope = 0, intercept = 0))
                #labs(title = "Heteroscedasticity", subtitle = "Check trends not captured by model")
        
        b <- ggplot(df, aes(residuals))+
                geom_histogram(bins = bins)
        
        c <- ggplot(df, aes(sample = residuals))+
                geom_qq()+
                geom_qq_line()#+
                #labs(title = "Quantile-quantile plot", subtitle = "check normality of residuals")
        
        d <- ggplot(df, aes(observations, fitted))+
                geom_point()+
                geom_abline()#+
                #labs(subtitle = "Check accuracy of model across whole dataset")
        
        return( (a + b) / (c + d) + plot_annotation(tag_levels = "a"))
} 


```

```{r data import}
data <- read_csv("Barbara_full_dataset_Bangor.csv")

flow <- read_csv("flow.csv") 
```

```{r clean}
long <- data|>
  select(-contains("RSV"), -contains("phi6"))|>
  group_by(sample_site_name, timestamp_sample_collected, timestamp_sample_received)|>
  mutate(biorep = row_number())|>
  pivot_longer(cols = contains("_gc_l"), names_to = "con_vir", values_to = "gc_l")|>
  separate(con_vir, into = c("method", "virus"), sep = "_", extra = "drop")|>
  filter(method != "SN")

long_sc <- long|>
  ungroup()|>
  mutate(gc_l_log = log10(gc_l),
         sample_ph_pre_ansis = scale(sample_ph_pre_ansis)[,1], 
         conductivity_ms_cm = scale(conductivity_ms_cm)[,1],
         ammonia_mg_l = scale(ammonia_mg_l)[,1],
         ophosph_mg_l = scale(ophosph_mg_l)[,1],
         Turbidity = scale(Turbidity)[,1])
  
diff <- long_sc|>
  select(-gc_l_log)|>
  pivot_wider(names_from = "method", values_from = "gc_l")|>
  group_by(virus)|>
  mutate(
      PEG_divby_AM = PEG / AM,
        AM_divby_PEG = AM / PEG)|>
  filter(PEG != 0, AM != 0,
         PEG_divby_AM < 10 & AM_divby_PEG < 10)

```

<br>

### Statistical methodology

Statistical tests were carried out in R (R Core Team, 2020); the full script and data are provided in a dedicated repository (<https://github.com/CameronPellett/Chem-Con-Bangor-Bath>).

Viral recovery with the amicon and BE-PEG methods was compared for six viruses (crAssphage, enterovirus, enterovirus D68, SARS-CoV-2, NoVGI, and NoVGII). The recovery in gene copies per litre was log~10~ transformed to meet assumptions for parametric analysis, then statistical comparisons were made using paired t-tests. Residual qq plots can be found in supplementary figure S1. 

Concentrations of ammonium, conductivity and orthophosphate have been shown to correlate with human populations, and thus relate to flow over short periods of time flow (REF). Therefore, the variability in flow had to be removed to assess the comparative effect of water chemistry variables, including ammonium, conductivity and orthophosphate, on each concentration method. To remove this variability, the gene copies per litre recovered by the amicon method was divided by the gene copies per litre recovered by the BE-PEG method. As each sample was paired, both sample methods were influenced by the same variability in flow, which was removed through this division. A linear mixed effects model was then used to assess the comparative effect of wastewater chemistry and quality variables (turbidity, pH, orthophosphate, conductivity, and ammonium) on the proportion of each concentration methods viral recovery, utilising random effects for the target virus to borrow strength in a combined assessment of all viruses (eq. 2). Prior to the model being fit, the dependent variable was power transformed using the Box-Cox method (eq. 1; eq. 2; Box and Cox, 1964), with lambda ($\lambda$) selected by maximising the log-Likelihood of a multiple linear regression with all predictors used in the final linear mixed effects model. The residual plots of the final model can be found in supplementary figure S2.



$$  
y(\lambda) = \{\frac{y ^ \lambda -1}{\lambda}, \quad if \space \lambda \neq 0;\quad log \space y, \quad if \space \lambda = 0 \rbrace  \tag{eq. 1}
$$

Where:
$y$ is the variable to be power transformed; and $\lambda$ is selected through maximising the log-Likelihood of a multiple linear regression with all predictors used in the final linear mixed effects model (eq . 2).


<br>


$$
y(\lambda)_{ij} = \beta_{0} + \beta_{1} \space turbidity_{ij} + \beta_{2} \space pH_{i} + \beta_{3} \space orthophosphate_{ij} + \\ \quad \quad \quad  \beta_{4} \space conductivity_{ij} + \beta_{5} \space ammonium_{ij} + \gamma_{j} \space virus_{ij} + \epsilon  \tag{eq. 2}
$$

Where:
$y(\lambda)_{ij}$ is $i$th value of $N$ total observations in the response variable after power transformation. $j$ relates to group of $F$ groups equal to the number of levels in the target virus variable;
 $\beta_{0}$ is the intercept, and $\beta_{1},...,\beta_{5}$ are the fixed-effects coefficient estimates; $\gamma_{j}$ is the random-effects coefficients in group $j$; and $\epsilon$ is the error between the model prediction and observation $i$ in group $j$.
 
<br>

## 3. Results

```{r method comparison}
meth_comp_data <- semi_join(long_sc, diff,
                            by = c("sample_site_name",
                                   "timestamp_sample_collected",
                                   "timestamp_sample_received", "biorep", "virus"))

qq_meth_comp <- meth_comp_data|>
  group_by(virus)|>
  mutate(row = row_number())|>
  ggplot(aes(sample = log10(gc_l)))+
  geom_qq()+
  geom_qq_line()+
  facet_grid(method ~ virus, scales = "free")

viruses_t.test <- meth_comp_data|>
  group_by(virus)|>
  summarise(max = max(gc_l, na.rm = T),
            min = min(gc_l, na.rm = T),
            mean = mean(gc_l, na.rm = T))|>
  bind_cols(stat = NA_character_,
            row = 1)

for (i in 1:nrow(viruses_t.test )) {
  filt_virus <- viruses_t.test [i,1][[1]]
  
  viruses_t.test [i,5] <- stat_paste_fun(t.test(gc_l_log ~ method, data = filter(meth_comp_data, virus == filt_virus), paired =TRUE), log = TRUE)
  
  
}

method_comp_plot <- meth_comp_data|>
  group_by(virus)|>
  mutate(row = row_number())|>
  left_join(viruses_t.test, by = c("virus", "row"))|>
  mutate(method = str_replace_all(method, c("PEG" = "BE-PEG")),
         virus = str_replace_all(virus, c("N1" = "SARS",
                                          "CrAss" = "crAss")))|>
  ggplot(aes(method, gc_l))+
  geom_boxplot(outlier.shape = NA)+
  geom_beeswarm(cex = 1.5, size = 1.5, alpha = 0.8)+
  geom_beeswarm(cex = 1.5, colour = "white", shape = ".")+
  geom_text(aes(label = stat, x = 1.5, y = max + (0.07 * (max - min))), size = 2.5)+
  facet_wrap(~virus, scale = "free")

```

### 3.1 Amicon and BE-PEG concentration methods have varied recovery depending on the target virus


The BE-PEG and amicon methods had varying performance depending on the target virus (Figure 1). Median gene copies per litre were similar between methods when targeting crAssphage 
(
AM: `` `r
meth_comp_data|>filter(virus == "CrAss", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` `` 
`r meth_comp_data|>filter(virus == "CrAss", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)`
gc/l;
BE-PEG: `` `r
meth_comp_data|>filter(virus == "CrAss", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "CrAss", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
),
enterovirus 
(
AM: `` `r
meth_comp_data|>filter(virus == "EV", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "EV", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l; BE-PEG: `` `r
meth_comp_data|>filter(virus == "EV", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "EV", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
)
and enterovirus D68 
(
AM: `` `r
meth_comp_data|>filter(virus == "EVD68", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "EVD68", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l;
BE-PEG: `` `r
meth_comp_data|>filter(virus == "EVD68", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "EVD68", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
),
and mean log~10~ transformed recoveries were not found to differ significantly between methods (paired t-test; p-value > 0.05). Conversely, the amicon method recovered greater median gene copies per litre when targeting SARS-CoV-2 
(
AM: `` `r
meth_comp_data|>filter(virus == "N1", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "N1", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l;
BE-PEG: `` `r
meth_comp_data|>filter(virus == "N1", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "N1", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
)
and norovirus GI 
(
AM: `` `r
meth_comp_data|>filter(virus == "NoVGI", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "NoVGI", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l; BE-PEG: `` `r
meth_comp_data|>filter(virus == "NoVGI", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "NoVGI", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
),
whilst the BE-PEG method recovered a greater median gene copies per litre when targeting norovirus GII  compared to the amicon method 
(
AM: `` `r
meth_comp_data|>filter(virus == "NoVGII", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "NoVGII", method == "AM")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l;
BE-PEG: `` `r
meth_comp_data|>filter(virus == "NoVGII", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` ``
`r meth_comp_data|>filter(virus == "NoVGII", method == "PEG")|>pull(gc_l)|>median()|>format(digits = 2, scientific = TRUE)` gc/l
).
Differences that were found to be significant when comparing mean log~10~ transformed gene copies per litre (paired t-test; p-value < 0.001). These conflicting results suggest neither method is consistently better than the other. Although, the amicon method may be best for projects which plan to primarily target SARS-CoV-2, particularly in catchments where low concentrations of virus in wastewater are frequent. 

<br>


```{r figure 1}
method_comp_plot+
  labs(y = "Gene copies per litre",
       x = "Method")
```

Figure 1: Comparison of viral recovery for amicon (AM) and beef extract polyethene glycol (BE-PEG) concentration methods. Statistical comparisons were made using paired t-tests after log transformation of the gene copies per litre. Recovery of crAssphage (crAss), enterovirus (EV), and enterovirus D68 (EVD68) could not be assumed to have differing means; whilst SARS-CoV-2 (SARS) and norovirus GI (NoVGI) had significantly greater mean recovery with the amicon method, and norovirus GII (NoVGII) had significantly greater mean recovery with the BE-PEG method.


<br>


```{r AM div PEG, include = F}
#boxcox2 <- MASS::boxcox(AM_divby_PEG ~ virus + sample_ph_pre_ansis + ammonia_mg_l + ophosph_mg_l + conductivity_ms_cm + Turbidity, 
 #                      data = diff, plotit = FALSE)

#this looks like chaos: essentially it runs the boxcox then selects the maximum log-likilihood (y) and its coresponding lambda (x)
boxcoxlambda2 <- MASS::boxcox(AM_divby_PEG ~ virus + sample_ph_pre_ansis + ammonia_mg_l + ophosph_mg_l + conductivity_ms_cm + Turbidity, 
                       data = diff)$x[which.max(MASS::boxcox(AM_divby_PEG ~ virus + sample_ph_pre_ansis + ammonia_mg_l + ophosph_mg_l + conductivity_ms_cm + Turbidity, 
                                                             data = diff)$y)]

boxcox_trans <- function(x, lambda){
  (x ^ lambda - 1) / lambda
}


lmm_bc2 <- lmer(AM_divby_PEG ~ (1|virus) + 
                 sample_ph_pre_ansis + ammonia_mg_l + ophosph_mg_l + conductivity_ms_cm + Turbidity,
               data = mutate(diff, AM_divby_PEG = boxcox_trans(AM_divby_PEG, boxcoxlambda2)))

lmm_residual_plot <- plot_assumptions(lmm_bc2,
                 mutate(diff, AM_divby_PEG = boxcox_trans(AM_divby_PEG, boxcoxlambda2))|>
                   filter(!is.na(AM_divby_PEG))|>
                   pull(AM_divby_PEG))



lmm_plot <- as_tibble(summary(lmm_bc2)[["coefficients"]], rownames = "variables")|>
  mutate(variables = str_replace_all(variables, c("methodPEG" = "PEG",
                                                  "methodAM" = "AM",
                                                  "sample_ph_pre_ansis" = "pH",
                                                  "ammonia_mg_l" = "Ammonium",
                                                  "ophosph_mg_l" = "Orthophosphate",
                                                  "conductivity_ms_cm" = "Conductivity")),
         #CI90 = (`Std. Error`*qt(1 - 0.1 / 2,  (round(348/2)) - 1)),
         CI95 = (`Std. Error`*qt(1 - 0.05 / 2,  (round(348/2)) - 1)),
         CI99 = (`Std. Error`*qt(1 - 0.01 / 2,  (round(348/2)) - 1)),
         CI999 = (`Std. Error`*qt(1 - 0.001 / 2,  (round(348/2)) - 1)),
         #label999 = c("99.9% CI", rep(NA_character_, 5)),
         #label99 = c("99%", rep(NA_character_, 5)),
         label95 = c( "95% CI", rep(NA_character_, 5)),
         #label90 = c( "90%", rep(NA_character_, 5)),
         signif = case_when((Estimate - CI999 > 0) | (Estimate + CI999 < 0) ~ "p-value < 0.001",
                            (Estimate - CI99 > 0) | (Estimate + CI99 < 0) ~ "p-value < 0.01",
                            (Estimate - CI95 > 0) | (Estimate + CI95 < 0) ~ "p-value < 0.05",
                            TRUE ~ "p-value > 0.05"))|>
  #filter(!str_detect(variables, "ntercept"))|>
  ggplot(aes( Estimate, variables))+
  geom_point(shape = 16, fill = "red")+
  #geom_errorbar(aes(xmin = Estimate - CI999,
   #                 xmax = Estimate + CI999), width = 0.2)+
  #geom_errorbar(aes(xmin = Estimate - CI99,
  #                  xmax = Estimate + CI99), colour = "grey40", width = 0.2)+
  geom_errorbar(aes(xmin = Estimate - CI95,
                    xmax = Estimate + CI95), width = 0.2)+#, colour = "grey50"
  geom_text_repel(aes(x = (Estimate), label = signif), nudge_y = 0.2, min.segment.length = 1, size = 2.5)+
  #geom_text_repel(aes(x = (Estimate + CI999), label = label999), nudge_y = 0.2)+
  #geom_text_repel(aes(x = (Estimate + CI99), label = label99), nudge_y = 0.2, colour = "grey40")+
  geom_text_repel(aes(x = (Estimate + CI95), label = label95), nudge_y = 0.2, size = 2.5)+#, colour = "grey50"
  geom_vline(xintercept = 0, linetype = 2)
```

### 3.2 The amicon method performes comparitively better than BE-PEG as concentrations of ammonium increase


A linear mixed effects model (lmm) was selected to assess the comparative effect of chemistry on viral recovery using both BE-PEG and amicon concentration methods. To remove the variability of other factors, including flow, the dependent variable was calculated as the proportion of viral recovery with the amicon method compared to the BE-PEG method (amicon recovery divided by BE-PEG recovery). The assessed proportion was then power transformed using the Box-Cox method to meet assumptions of a linear regression, selecting the lambda that maximised the log-Likelihood of a multiple linear model using all predictors used in the final lmm
(
lambda = `` `r
round(boxcoxlambda2, 2)` ``
`r round(boxcoxlambda2, 2)`
).
The lmm found ammonium had a significant positive effect (p-value < 0.01), suggesting the amicon method performs better when there are greater concentrations of inhibitors. However, none of the other chemistry variables had significantly different effects on the amicon and BE-PEG methods (p-value > 0.05), and the intercept was not significant suggesting the methods did not have different mean recoveries when assessing all viruses. These results suggest the amicon method may be preferable when catchments have high concentrations of ammonium, including those with high levels of agricultural runoff in the sewage network.


<br>


```{r figure 2, fig.height = 5}
lmm_plot+
  theme(axis.title.y = element_blank())

```

Figure 2: Linear mixed model assessing the effect of chemistry on the proportion of amicon BE-PEG gene copies per litre recovery (amicon recovery divided by BE-PEG recovery). The assessed proportion was power transformed using the Box-Cox method, selecting the lambda that maximised the log-Likelihood 
(
lambda = `` `r
round(boxcoxlambda2, 2)` ``
`r round(boxcoxlambda2, 2)`
).
The target virus was included as random effects to account for the variability between viruses, borrowing strength to assess the effect of chemistry. Estimates less than zero suggest the amicon method performs worse than the BE-PEG method when predictor variables increase, or on average with regards to the intercept. Estimates are labeled with their p-values and the line ranges around the estimate indicate the 95% confidence intervals (CI).  

<br>

## References 

Box, G.E. and Cox, D.R., 1964. An analysis of transformations. Journal of the Royal Statistical Society: Series B (Methodological), 26(2), pp.211-243.

<br>

## Supplementary 

<br>


```{r S1}
qq_meth_comp+
  ylab("log10 recovery (gc/l)")
```

Figure S1:quantile-quantile normality of the log10 transformed recovery of each virus with each concentration method.

<br>

```{r S2}
lmm_residual_plot
```

Figure S2: Residual plot of the linear mixed effects model. Residual plots of (a) homoscedasticity, (b) histogram normality, (c) quantile-quantile normality with x as a theoretical normal distribution, and (d) true observations against fitted values.
